The State of Urbit\\Eight Years After the Whitepaper

### 2024-04-09 mss.tex
Post

The other big change is the public key infrastructure ({\sc pki}).  The whitepaper's {\sc pki}, which was live from 2013 to 2019 (TODO: check date), had no double-spend protection and was effectively centralized.  To fix this, the {\sc pki} was moved to a set of Ethereum contracts.  Going on-chain made the ownership ledger for Urbit IDs censorship-resistant ... until a year later when high transaction fees on Ethereum brought network growth to a standstill and the core developers were forced to write a custom "naive rollup" Ethereum contract.  This brought the price of spawning a new Urbit node down from hundreds of dollars to under a dollar, at the cost of on-chain interoperability.

Going on-chain also enabled "personal continuity breaches", where a ship owner can notify the network through the chain that their Urbit node was rebooted from scratch, due to data loss, unrecoverable error, or compromised security.  This feature necessitated some rework of the kernel to ensure data consistency across vanes when receiving a notification that a peer had breached.

@@

A major part of making the namespace more usable was to expose namespace queries over the network.  Designed in 2021 and deployed in 2023, the "Fine" remote scry network protocol was added to Urbit.  The first use cases were Clay, which uses it for downloading desks, and the Gall `%grow` namespace, described in more detail in a later section.

This protocol lets one ship fetch data at a namespace path bound by another ship.  The runtime includes a cache so later requests for the same data are typically served from the runtime cache rather than requiring Arvo's `+peek` arm to be called again.  In addition to the scalability benefits this provided, it also represented a fairly large conceptual shift: instead of Urbit's networking only being able to express commands (writes) over the network, it could now also support reads.

@@

This can be band-aided by baking the specifics of the interaction into Jael, so it knows to tell Ames about the breach before Gall.  This approach works as long as the data dependencies among modules remains acyclic, but as soon as two modules could both send moves to each other upon being notified – meaning they need to be notified "at the same time" – it becomes nontrivial to provide consistency.  It could be achieved through a two-phase commit system where Jael delivers two moves for each notification, and the receiving modules would know to update their state in response to the first notification but only emit moves upon hearing the second notification.

@@

The latest "shrubbery" experiment uses a hybrid move ordering: depth-first for local pokes within the same application, and breadth-first otherwise.

@@

The latest "shrubbery" experiment attempts a hybrid solution to get the best of both modalities.  A typical shrubbery "app" will be broken into a hierarchy of event handlers at various subpaths.  When sending a move "downward" into a subpath, the system will handle that move depth-first, since an app is presumed to know the characteristics of its sub-apps.  These moves will not need to be acknowledged, since they will be truly depth-first.  When passing a move "upward" or "sideways", i.e. outside of the sub-tree, that move will be appended to one global queue, to be handled breadth-first.

@@

The capability to perform an ``individual continuity breach'' was added.  Before going on-chain, if an Urbit suffered any major data loss, it would be permanently dead and could never communicate with peers again.  An individual continuity breach allows a user to post a transaction to chain that increments the "rift" number for their address, signaling to the network that the ship has been restarted from scratch.  Peer ships that were storing message sequence numbers for the breached ship reset their networking state for the breached ship.

@@

It should be noted that galactic governance is purely voluntary: because users have full control over their own personal servers, they retain the power to "vote with their feet" and decide to use a different {\sc pki} or a different governance structure.

@@

In 2020, Ethereum transaction fees rose to levels high enough that Urbit network growth ground to a halt.  The core developers, after evaluating the nascent "zero-knowledge" and "optimistic" rollup technologies for reducing fees, determined they were not yet production-worthy and wrote a simpler rollup that they called a "naive rollup".  This brought the price of spawning a planet down from roughly \$100 to under \$1.

@@

Each ship's Azimuth state is stored either in the rollup or in the original contract.  Galaxies cannot be moved into the rollup.  For now, addresses cannot be moved back out of the rollup.

\subsection{New Ames & New Jael (2019)}

@@

The ``new'' Ames and Jael – now in their fourth year of use, with Ames having had some refactoring and extension – support individual continuity breaches and on-chain sponsorship changes ("escapes").

@@

In early 2023, the responsibility for core development transferred from the Tlon Corporation, which had managed it since 2013, to the newer Urbit Foundation, a nonprofit.  UF adopted and continued to modify Tlon's release workflow, including CI/CD (continuous integration / continuous deployment), multi-stage canary deployments, and repository management practices.  This process has more recently been extended to include a liaison period for each Kelvin update where organizations that use Urbit can use the prerelease software on their own infrastructure first and report any bugs to UF.

@@

The list goes on.  While Urbit still has flaws and occasional crashes, its level of robustness is strikingly higher than it was when the whitepaper was written.  By early 2025, we expect Urbit to be reliable enough to pass a security audit that includes penetration testing.

@@

The runtime has absorbed a large percentage of the work that has been done in the system in the last eight years.  The naive tree-walking Nock interpreter was replaced by a custom bytecode interpreter.  The system was made to recover from out-of-memory errors and automatically reclaim memory.  A memory allocator trick called "pointer compression" was used to increase data storage from 2GB to 8GB while remaining a 32-bit interpreter.  Demand paging was introduced, reducing the cost of running an Urbit ship from over \$10 per month to \$0.15 per month on some hosting platforms.

The runtime was broken into two processes, allowing I/O and Arvo event processing to run in parallel.  The event log was moved out of a custom file into an {\sc lmdb} database, then broken up into "epochs" allowing for event log rollover and truncation.  A lot of work has gone into the I/O drivers in the runtime, including full rewrites of the {\sc http}, Ames, and terminal drivers.

@@

TODO: describe Clay tombstoning somewhere

@@

Having many copies of the same library in memory is a well-known problem with static linking, which is logically what Ford does to link Hoon programs together.  The linking step is a trivial dynamically typed (``vase-mode'') `cons' operation, since Hoon's variable lookup simply traverses the "subject" (environment) type left-to-right.  Importing a library means prepending that library to the environment.

@@

@@

As users install more applications, it becomes increasingly important for the system to be able to protect itself and other apps from a malicious app.  A basic measure was introduced in 2023 so that a Gall agent can tell which other local agent sent it a command.  This "poke provenance" was enough to prevent certain kinds of confused deputy attacks involving ships delegating some functionality to their moons.

For full protection, however, a system of "userspace permissions" is required, where the user must approve an agent's access to kernel features and other applications with some granularity.  Such a system was prototyped in late 2022 and is scheduled for further development and deployment later in 2024.

@@

TODO: decide whether to have one – could try to take the first part of the appendix, perhaps

An appendix to this discussion addresses developments from the whitepaper by section.

\tombstone{}

@@

TODO: organize better

Accessing Urbit's namespace as a {\sc fuse} filesystem still has not been implemented.  Some of the whitepaper was written in the "prophetic present tense", and while a lot of those prophecies have come true, that one remains aspirational.

@@

Urbit is still "neither secure nor reliable", but it is much more reliable than back then.  I estimate about two more "nines" have been added to its uptime percentage.  Error handling has improved dramatically.  Userspace apps can be managed by a user.  The runtime and kernel both clean up after themselves better in several different ways, so crashes due to resource exhaustion are much less common.

All of the commands that users now lean on for managing resource usage were written after the whitepaper:
- `|trim` cache pruning
- `|pack` defragmentation
- `|meld` deduplication
- `chop` event log truncation
- `roll` event log rollover

Nock's changes going from 5K to 4K resulted from `~fodwyt-ragful`'s bytecode interpreter in 2018.  There were three changes.

The Nock 5 equality check opcode was tweaked to enforce taking two subexpressions, rather than optionally taking a single expression that could return a cell (pair), whose head and tail would then be compared for equality.  This optionality turned out to be annoying when writing an interpreter, and it provided little utility.

Nock version 4 added a new Nock 10 opcode to replace part of a tree with a new value.  Creating mutant trees like this is a basic operation that Hoon code performs regularly, using the `%=` rune (TODO: should that be centar?) and its variants (`%*`, `%_`, `=.`, and `=:`).  Before the addition of the Nock 10 opcode, the Hoon compiler would use the type of the trees involved in a mutation expression to codegen a tuple of Nock "auto-cons" expressions (cells of expressions that return cells of their results) that would reconstruct the mutant tree out of the new subtree cons'd together with the remaining pieces of the original tree.

@@

A tertiary benefit to adding Nock 10 was that the calling convention for "slamming a gate" (providing an argument to an anonymous function and running the function's body) could now be expressed in statically generated Nock code without needing to invoke the Hoon compiler.  This allowed a bit of additional expressivity in some Hoon standard library routines that call into untyped Nock.

Hoon has grown, unfortunately.  There have been two or three revisions of syntax and type system.  The latest change was adding "doccords", a form of docstring, to the syntax.

Parsing and compiling are still slow, unfortunately, but there have been massive performance improvements to Hoon compilation, especially in memory usage.  In 2017 compiling kernelspace would use around a gigabyte of {\sc ram}; now it uses on the order of a hundred megabytes.  Reducing this memory pressure alleviated a lot of crashes from out-of-memory errors ("bail:meme" errors) – it was a major contributor to the increase in overall reliability.

@@

Hoon's recent changes have reduced the prevalence of the "TMI problem".  Since Hoon's type system supports subtyping, it is common for code to fail to compile due to having "too much information" (TMI) – sometimes a more specific type cannot be used in place of a more general one, namely when that type is in "contravariant position".

For example, if the type of some field in the Hoon subject was a list but got specialized in a conditional branch to be a non-null list, attempting to "overwrite" (really, create a mutant copy of) that field in the subject with a normal list, the compiler will throw a type error because it cannot guarantee that the normal list is not null.  If the compiler did not know the subject's list field was non-null, then overwriting it with some other list would just.work.  Almost always, the eventual result of the expression admits a null list, so the type error is an annoyance rather than an assurance.

Hoon type system changes have also included multiple attempts at resolving another difficulty with subtyping: "wetness".  A "wet core" is a Hoon core (pair of code and data, where the code is run using the whole core as the subject) with a "sample" (argument slot – the head of the "data" tail by convention in the type system) whose type can vary.  This is very closely related to the mathematical idea of parametric polymorphism.  Like in many languages (including Haskell), Hoon does not strictly guarantee "parametricity", but wetness is used to implement parametrically polymorphic functions and even type constructors (called "mold builders" in Hoon).

Wetness has an issue where the "faces" (names for particular slots within the tree) in the sample type in the definition of the wet core can differ from the faces in the sample type at any given call site.  Deciding which of those faces to use in the body of the code is a subtle, thorny problem.  The Hoon at the time of the whitepaper had one solution; another was written afterward, around 2017; a third was deployed in 2019.  This last one, called "repainting", is generally thought to be the right strategy, but its implementation remains unfinished in 2024 – it has an issue where the precedence of call-site vs. definition-site face names is not preserved under recursion.  This issue will need to be fixed before we can say the right answer to this issue has been achieved.

Wetness in general is an interesting and novel approach to both polymorphic and recursive types, both of which are directly represented as lazily evaluated types (Urbit core devs often call the "evaluation" step "type inference" informally, but it corresponds to the "type synthesis" direction of a bidirectional typechecker in type theory terminology).  This has pros and cons.

@@

Arvo's "duct" as a reified vane call stack is still very much in use.  The patterns of use have evolved somewhat.  It is now much more common for vanes to use "synthetic ducts" as the origins of certain operations to disconnect them from their cause.

For example, Ames uses a synthetic duct for each "flow" (Ames's version of a socket connection) to set packet re-send timers for that flow and pass requests to other vanes.  This disconnects the flow's duct from the duct that the runtime used to inject the event into Arvo, preventing Unix process restarts from causing desynchronization.

@@

The whitepaper says "most data is now in Clay".  This is no longer true.  Now most data lives in Gall, in agents.  Clay has proven difficult to use for storing application data for a number of reasons, and Gall has become more featureful.  There is consensus among the core devs that a better userspace model should be devised to unify Clay and Gall.  One such model, called "shrubbery", is now being investigated.

Urbit is still a non-preemptive OS, but it could be made preemptive by letting the kernel emit a Nock hint to the runtime to enforce a time limit on the enclosed computation.  If it failed to complete within the given time limit, the runtime would abort the entire Arvo event.  This would be considered a nondeterministic error ("bail:fail"), like a user-initiated event interruption (Ctrl-C).  This could potentially even be used to turn Urbit into a real-time OS.

## Section 2.4 Philosophy of Ames Networking

The philosophy behind the Ames networking system is still accurate despite multiple Ames rewrites and the move onto a blockchain.

## Section 3.1 Uniform Persistence

This section is still true.  A lot of work has gone into improving the runtime's implementation of uniform persistence:
- demand paging to allow storage of more data on disk than fits in {\sc ram}
- a guard page to efficiently prevent stack smashing
- a snapshot migration framework to facilitate upgrades
- separating ephemeral and persistent state to avoid extraneous disk writes
- using the {\sc lmdb} database for the event log, instead of a custom binary file
- the Ares project's Persistent Memory Arena ({\sc pma}), a general-purpose single-level store that can efficiently manage the persistence and paging for 64 terabytes of data, using a copy-on-write B+ tree
- the "epoch system", which broke the event log into epochs, allowing event log rollover and truncation
- persistent Nock caches, allowing arbitrary computations to be memoized in the runtime in a way that persists across Arvo events and Unix process restarts

## Section 3.2 "Source-Independent Networking"

Ames is still sequential and idempotent for commands.  Despite having been described as "content-centric" and "source-independent" in the whitepaper, those two properties are only now being realized, in the "directed messaging" project that is currently underway.  Directed messaging refactors Ames to be much more like the Named Data Networking project: packets and messages are both constrained to be request/response, with routing being achieved using a Pending Interest Table just like in NDN.

@@

This is only the latest in a series of changes that have been made to Ames since the whitepaper:
- it is now backed by Azimuth, on the blockchain
- it supports personal continuity breaches
- "stateless forwarding" brought orders of magnitude more scalability
- the \lstinline[style=inlinecode]{%clog} backpressure system was added to prevent unbounded packet queueing on publisher ships; this was later tuned for better performance
- flows can now be closed, so their memory resources can be reclaimed
- "dead flow consolidation": outbound requests to unresponsive ships now share one global timer every two minutes, dramatically reducing the number of Arvo events
- the \lstinline[style=inlinecode]{%flub} system was added to close a DoS vulnerability: the system can drop incoming requests until it is ready to process them, without any queueing
- the \lstinline[style=inlinecode]{|snub} and \lstinline[style=inlinecode]{|ruin} commands were added to allow manual blacklisting of peers
- network reads have been implemented
  - the Fine remote scry protocol
  - the \lstinline[style=inlinecode]{%chum} one-to-one encrypted remote scry protocol
  - the one-to-many encrypted remote scry protocol

## Section 3.4 Interruption, etc.

There are still no worker threads, but a potential strategy has been devised for them, and the Ares {\sc pma} should support parallel execution.

There are still no event timeouts, but we do not expect them to be difficult to implement.

## Section 3.5 Solid-State Interpreter

There is still no unikernel, but a spec has been written for one.  There is no Raft cluster yet, but the \lstinline[style=inlinecode]{%phoenix} app now uses encrypted remote scry to back up application state.  There have also been proposals to use FoundationDB to replicate the event log storage.

## 4.2 Larval State

@@

4.3 Step Model

This model is still in effect.

TODO Is \lstinline[style=inlinecode]{%lick} a step or a packet? ask ~mopfel

## 5.1 Nouns

The Murmur3 system was aspirational in the whitepaper, but was subsequently realized.  A few years ago, a horrendous bug was found and fixed in the ``mug'' short hash for cells:  the algorithm was symmetric, so \lstinline[style=inlinecode]{[1 0]} hashed to the same value as \lstinline[style=inlinecode]{[0 1]}.  This led to far more hash collisions than there should have been, degrading the asymptotics of a number of data structures.

## Section 6.3 Nock Optimization

The ``jet dashboard'' (the module in the runtime responsible for registering, matching, and running jets) has been rewritten multiple times since the whitepaper.  \patp{fodwyt-ragful} has a proposal for a jet dashboard that would match Nock formulas on a noun equality basis, rather than matching Nock cores (a code/data pair used as the unit of code organization) while ignoring mutable slots, the way the previous ones have worked.  This would have the advantage of simplicity, but it could be more difficult to implement more complex pieces of code, such as closures.

The Ares project has another new approach to jet matching, which is to resolve the matching statically during a transpilation step from Nock to a more optimized representation.  This is facilitated by a novel static analysis technique for Nock, called Subject Knowledge Analysis, which analyzes the data dependencies within a Nock formula, allowing inference about the tree shape of the ``subject'' (environment, or scope) at different points of the computation.

This should allow for more efficient registerization of intermediate values, and it also provides enough information to statically compute whether a jet will match.  The goal is that Ares will convert a hot loop in Nock into similar machine code that would be emitted by a C compiler, including direct jumps back to the start of the loop.

The Rust interpreter project mentioned in the whitepaper is dead, but in the meantime several other interpreters have been written:
- Jaque, written using the Graal/Truffle JIT framework
- NockJS, an interpreter in JavaScript that uses trampolining
- cl-urbit, a Common Lisp interpreter used as a jet dashboard testbed
- Ares, written in Rust and C, intended as the next mainline interpreter version

@@

## 6.5 Transition Function

The whitepaper mentions off-handedly that Nock could not be replaced.  While the author does not expect Nock to be changed, much less replaced, this could be done if needed.  The implementation is left as an exercise for the reader.

## 7.1 Hoon Semantics

>>>>

The other big change is the public key infrastructure ({\sc pki}).  The whitepaper's {\sc pki}, which was live from 2013 to 2019 (TODO: check date), had no double-spend protection and was effectively centralized.  To fix this, the {\sc pki} was moved to a set of Ethereum contracts.  Going on-chain made the ownership ledger for Urbit IDs censorship-resistant ... until a year later when high transaction fees on Ethereum brought network growth to a standstill and the core developers were forced to write a custom ``naive rollup'' Ethereum contract.  This brought the price of spawning a new Urbit node down from hundreds of dollars to under a dollar, at the cost of on-chain interoperability.

Going on-chain also enabled ``personal continuity breaches'', in which a ship owner can notify the network through the chain that their Urbit node was rebooted from scratch, due to data loss, unrecoverable error, or compromised security.  This feature necessitated some rework of the kernel to ensure data consistency across vanes when receiving a notification that a peer had breached.

@@

A major part of making the namespace more usable was to expose namespace queries over the network.  Designed in 2021 and deployed in 2023, the ``Fine''\footnote{Pronounced \emph{fee-NAY}.} remote scry network protocol was added to Urbit.  The first use cases were Clay, which uses it for downloading desks, and the Gall \lstinline[style=inlinecode]{%grow} namespace, described in more detail in a later section.

This protocol lets one ship fetch data at a namespace path bound by another ship.  The runtime includes a cache so later requests for the same data are typically served from the runtime cache rather than requiring Arvo's \lstinline[style=inlinecode]{+peek} arm to be called again.  In addition to the scalability benefits this provided, it also represented a fairly large conceptual shift: instead of Urbit's networking only being able to express commands (writes) over the network, it could now also support reads.

@@

This can be band-aided by baking the specifics of the interaction into Jael, so it knows to tell Ames about the breach before Gall.  This approach works as long as the data dependencies among modules remains acyclic, but as soon as two modules could both send moves to each other upon being notified – meaning they need to be notified ``at the same time'' – it becomes nontrivial to provide consistency.  It could be achieved through a two-phase commit system where Jael delivers two moves for each notification, and the receiving modules would know to update their state in response to the first notification but only emit moves upon hearing the second notification.

@@

The latest ``shrubbery'' experiment uses a hybrid move ordering: depth-first for local pokes within the same application, and breadth-first otherwise.

@@

The latest ``shrubbery'' experiment attempts a hybrid solution to get the best of both modalities.  A typical shrubbery ``app'' will be broken into a hierarchy of event handlers at various subpaths.  When sending a move ``downward'' into a subpath, the system will handle that move depth-first, since an app is presumed to know the characteristics of its sub-apps.  These moves will not need to be acknowledged, since they will be truly depth-first.  When passing a move ``upward'' or ``sideways'', i.e. outside of the sub-tree, that move will be appended to one global queue, to be handled breadth-first.

@@

The capability to perform an ``individual continuity breach'' was added.  Before going on-chain, if an Urbit suffered any major data loss, it would be permanently dead and could never communicate with peers again.  An individual continuity breach allows a user to post a transaction to chain that increments the ``rift'' number for their address, signaling to the network that the ship has been restarted from scratch.  Peer ships that were storing message sequence numbers for the breached ship reset their networking state for the breached ship.

@@

It should be noted that galactic governance is purely voluntary: because users have full control over their own personal servers, they retain the power to ``vote with their feet'' and decide to use a different {\sc pki} or a different governance structure.

@@

In 2020, Ethereum transaction fees rose to levels high enough that Urbit network growth ground to a halt.  The core developers, after evaluating then-nascent ``zero-knowledge'' and ``optimistic'' rollup technologies for reducing fees, determined they were not yet production-worthy and wrote a simpler rollup that they called a ``naive rollup''.  This brought the price of spawning a planet down from roughly \$100 to under \$1.

@@

Each ship's Azimuth state is stored either in the rollup or in the original contract.  Galaxies cannot be moved into the rollup.  For the time being, addresses cannot be moved back out of the rollup.

\subsection{New Ames \& New Jael (2019)}

@@

The ``new'' Ames and Jael – now in their fourth year of use, with Ames having had some refactoring and extension – support individual continuity breaches and on-chain sponsorship changes (``escapes'').

@@

In early 2023, the responsibility for core development transferred from the Tlon Corporation, which had managed it since 2013, to the newer Urbit Foundation, a nonprofit.  The Urbit Foundation has become the custodian of the Urbit kernel and ecosystem.  UF adopted and continued to modify Tlon's release workflow, including CI/CD (continuous integration/continuous deployment), multi-stage canary deployments, and repository management practices.  This process has more recently been extended to include a liaison period for each Kelvin update where organizations that use Urbit can use the prerelease software on their own infrastructure first and report any bugs to UF.

@@

The list goes on.  While Urbit still has flaws and occasional crashes, its level of robustness is strikingly higher than it was when the whitepaper was written.  By early 2025, we expect Urbit to be reliable enough to pass a security audit that includes penetration testing.  At that point, custodial cryptographic applications will be viable on the platform.

@@

The runtime has absorbed a large percentage of the work that has been done in the system in the last eight years.  The naive tree-walking Nock interpreter was replaced by a custom bytecode interpreter.  The system was made to recover from out-of-memory errors and automatically reclaim memory.  A memory allocator trick called ``pointer compression'' was used to increase data storage from 2~GB to 8~GB while remaining a 32-bit interpreter.  Demand paging was introduced, reducing the cost of running an Urbit ship from over \$10 per month to \$0.15 per month on some hosting platforms.

The runtime was broken into two processes, allowing I/O and Arvo event processing to run in parallel.  The event log was moved out of a custom file into an {\sc lmdb} database, then broken up into ``epochs'' allowing for event log rollover and truncation.  The I/O drivers in the runtime have continued to be developed, including full rewrites of the {\sc http}, Ames, and terminal drivers.

@@

@@

Having many copies of the same library in memory is a well-known problem with static linking, which is logically what Ford does to link Hoon programs together.  The linking step is a trivial dynamically typed (``vase-mode'') `cons' operation, since Hoon's variable lookup simply traverses the ``subject'' (environment) type left-to-right.  Importing a library means prepending that library to the environment.

@@

\subsection{Clay Tombstoning}

Besides noun deduplication, Clay has also introduced a tombstoning system.  Files in all revisions before the current one may be ``tombstoned'', meaning that the old data has been deleted and just the reference remains.  Tombstoning policy can be configured for files, directories, etc, so that large files aren't duplicated while full revision history can be retained for other things where it's appropriate.

Tombstoning is not at odds with the referential transparency of scries, since a scry attempt may return \lstinline[style=inlinecode]{[~ ~]}, a permanent scry failure (that resource will never be available).

@@

As users install more applications, it becomes increasingly important for the system to be able to protect itself and other apps from a malicious app.  A basic measure was introduced in 2023 so that a Gall agent can tell which other local agent sent it a command.  This ``poke provenance'' was enough to prevent certain kinds of confused deputy attacks involving ships delegating some functionality to their moons.

For full protection, however, a system of ``userspace permissions'' is required, where the user must approve an agent's access to kernel features and other applications with some granularity.  Such a system was prototyped in late 2022 and is scheduled for further development and deployment later in 2024.

@@

In conclusion, Urbit has grown to realize many of the promises of the 2016 whitepaper.  The system is more reliable, more scalable, more extensible, and more secure than it was then.  The network has grown to include thousands of users, and the system is now capable of supporting a much larger userbase.  While more work remains to be done, the kernel of Urbit is sound and the contours of the kelvin-cooled system have begun to assume more resolution.  \tombstone{}

@@

This appendix addresses developments from the whitepaper by section.  While this is somewhat narratively unsatisfactory, it provides an easy reference.  In cases where no significant changes have been made, the section has been elided.

Accessing Urbit's namespace as a {\sc fuse} filesystem still has not been implemented.  Some of the whitepaper was written in the ``prophetic present tense'', and while a lot of those prophecies have come true, that one remains aspirational.

@@

Urbit is still ``neither secure nor reliable'', but it is much more reliable than back then.  We estimate that about two more ``nines'' have been added to its uptime percentage.  Error handling has improved dramatically.  Userspace apps can be managed by a user.  The runtime and kernel both clean up after themselves better in several different ways, so crashes due to resource exhaustion are much less common.

All of the commands that users now lean on for managing resource usage were written after the whitepaper:
\begin{itemize}
  \item  \lstinline[style=inlinecode]{|trim} cache pruning
  \item  \lstinline[style=inlinecode]{|pack} defragmentation
  \item  \lstinline[style=inlinecode]{|meld} deduplication
  \item  \lstinline[style=inlinecode]{chop} event log truncation
  \item  \lstinline[style=inlinecode]{roll} event log rollover
\end{itemize}

Nock's changes going from 5K to 4K resulted from \patp{fodwyt-ragful}'s bytecode interpreter in 2018.  Three changes were made:

\begin{enumerate}
  \item  The Nock 5 equality check opcode was tweaked to enforce taking two subexpressions, rather than optionally taking a single expression that could return a cell (pair), whose head and tail would then be compared for equality.  This optionality turned out to be annoying when writing an interpreter, and it provided little utility.
  \item  Nock version 4 added a new Nock 10 opcode to replace part of a tree with a new value.  Creating mutant trees like this is a basic operation that Hoon code performs regularly, using the \lstinline[style=inlinecode]{%=} rune (TODO: should that be centar?) and its variants (\lstinline[style=inlinecode]{%*}, \lstinline[style=inlinecode]{%_}, \lstinline[style=inlinecode]{=.}, and \lstinline[style=inlinecode]{=:}).  Before the addition of the Nock 10 opcode, the Hoon compiler would use the type of the trees involved in a mutation expression to codegen a tuple of Nock ``auto-cons'' expressions (cells of expressions that return cells of their results) that would reconstruct the mutant tree out of the new subtree cons'd together with the remaining pieces of the original tree.

@@

  \item  A tertiary benefit to adding Nock 10 was that the calling convention for ``slamming a gate'' (providing an argument to an anonymous function and running the function's body) could now be expressed in statically generated Nock code without needing to invoke the Hoon compiler.  This allowed a bit of additional expressivity in some Hoon standard library routines that call into untyped Nock.
\end{enumerate}

Parsing and compiling are still slow, unfortunately, but there have been massive performance improvements to Hoon compilation, especially in memory usage.  In 2017 compiling kernelspace would use around a gigabyte of {\sc ram}; now it uses on the order of a hundred megabytes.  Reducing this memory pressure alleviated a lot of crashes from out-of-memory errors (\lstinline[style=inlinecode]{bail:meme} errors) – it was a major contributor to the increase in overall reliability.

@@

Hoon's recent changes have reduced the prevalence of the ``{\sc tmi} problem''.  Since Hoon's type system supports subtyping, it is common for code to fail to compile due to having ``too much information'' ({\sc tmi}) – sometimes a more specific type cannot be used in place of a more general one, namely when that type is in ``contravariant position''.  For example, if the type of some field in the Hoon subject was a list but got specialized in a conditional branch to be a non-null list, attempting to ``overwrite'' (really, create a mutant copy of) that field in the subject with a normal list, the compiler will throw a type error because it cannot guarantee that the normal list is not null.  If the compiler did not know the subject's list field was non-null, then overwriting it with some other list would just.work.  Almost always, the eventual result of the expression admits a null list, so the type error is an annoyance rather than an assurance.

Hoon type system changes have also included multiple attempts at resolving another difficulty with subtyping: ``wetness''.  A ``wet core'' is a Hoon core (pair of code and data, where the code is run using the whole core as the subject) with a ``sample'' (argument slot – the head of the ``data'' tail by convention in the type system) whose type can vary.  This is very closely related to the mathematical idea of parametric polymorphism.  Like in many languages (including Haskell), Hoon does not strictly guarantee ``parametricity'', but wetness is used to implement parametrically polymorphic functions and even type constructors (called ``mold builders'' in Hoon).

Wetness has an issue where the ``faces'' (names for particular slots within the tree) in the sample type in the definition of the wet core can differ from the faces in the sample type at any given call site.  Deciding which of those faces to use in the body of the code is a subtle, thorny problem.  The Hoon at the time of the whitepaper had one solution; another was written afterward, around 2017; a third was deployed in 2019.  This last one, called ``repainting'', is generally thought to be the right strategy, but its implementation remains unfinished in 2024 – it has an issue where the precedence of call-site vs. definition-site face names is not preserved under recursion.  This issue will need to be fixed before we can say the right answer to this issue has been achieved.

Wetness in general is an interesting and novel approach to both polymorphic and recursive types, both of which are directly represented as lazily evaluated types (Urbit core devs often call the ``evaluation'' step ``type inference'' informally, but it corresponds to the ``type synthesis'' direction of a bidirectional typechecker in type theory terminology).  This has pros and cons.

@@

Arvo's ``duct'' as a reified vane call stack is still very much in use.  The patterns of use have evolved somewhat.  It is now much more common for vanes to use ``synthetic ducts'' as the origins of certain operations to disconnect them from their cause.

For example, Ames uses a synthetic duct for each ``flow'' (Ames's version of a socket connection) to set packet re-send timers for that flow and pass requests to other vanes.  This disconnects the flow's duct from the duct that the runtime used to inject the event into Arvo, preventing Unix process restarts from causing desynchronization.

@@

The whitepaper says ``most data is now in Clay''.  This is no longer true.  Now most data lives in Gall, in agents.  Clay has proven difficult to use for storing application data for a number of reasons, and Gall has become more featureful.  There is consensus among the core devs that a better userspace model should be devised to unify Clay and Gall.  One such model, called ``shrubbery'', is now being investigated.

Urbit is still a non-preemptive OS, but it could be made preemptive by letting the kernel emit a Nock hint to the runtime to enforce a time limit on the enclosed computation.  If it failed to complete within the given time limit, the runtime would abort the entire Arvo event.  This would be considered a nondeterministic error (\lstinline[style=inlinecode]{bail:fail}), like a user-initiated event interruption (Ctrl-C).  This could potentially even be used to turn Urbit into a real-time OS.

\paragraph{3.1 Uniform Persistence}

Significant work has gone into improving the runtime's implementation of uniform persistence:
\begin{itemize}
  \item  demand paging to allow storage of more data on disk than fits in {\sc ram}
  \item  a guard page to efficiently prevent stack smashing
  \item  a snapshot migration framework to facilitate upgrades
  \item  separating ephemeral and persistent state to avoid extraneous disk writes
  \item  using the {\sc lmdb} database for the event log, instead of a custom binary file
  \item  the Ares project's Persistent Memory Arena ({\sc pma}), a general-purpose single-level store that can efficiently manage the persistence and paging for 64 terabytes of data, using a copy-on-write B+ tree
  \item  the ``epoch system'', which broke the event log into epochs, allowing event log rollover and truncation
  \item  persistent Nock caches, allowing arbitrary computations to be memoized in the runtime in a way that persists across Arvo events and Unix process restarts
\end{itemize}

\paragraph{3.2 Source-Independent Networking}

Ames is still sequential and idempotent for commands.  Despite having been described as ``content-centric'' and ``source-independent'' in the whitepaper, those two properties are only now being realized, in the ``directed messaging'' project that is currently underway.  Directed messaging refactors Ames to be much more like the Named Data Networking project: packets and messages are both constrained to be request/response, with routing being achieved using a Pending Interest Table just like in {\sc ndn}.

@@

This is only the latest in a series of changes that have been made to Ames since the whitepaper:
\begin{itemize}
  \item  it is now backed by Azimuth, on the blockchain
  \item  it supports personal continuity breaches
  \item  ``stateless forwarding'' brought orders of magnitude more scalability
  \item  the \lstinline[style=inlinecode]{%clog} backpressure system was added to prevent unbounded packet queueing on publisher ships; this was later tuned for better performance
  \item  flows can now be closed, so their memory resources can be reclaimed
  \item  ``dead flow consolidation'': outbound requests to unresponsive ships now share one global timer every two minutes, dramatically reducing the number of Arvo events
  \item  the \lstinline[style=inlinecode]{%flub} system was added to close a DoS vulnerability: the system can drop incoming requests until it is ready to process them, without any queueing
  \item  the \lstinline[style=inlinecode]{|snub} and \lstinline[style=inlinecode]{|ruin} commands were added to allow manual blacklisting of peers
  \item  network reads have been implemented
  \begin{itemize}
    \item  the Fine remote scry protocol
    \item  the \lstinline[style=inlinecode]{%chum} one-to-one encrypted remote scry protocol
    \item  the one-to-many encrypted remote scry protocol
  \end{itemize}
\end{itemize}

\paragraph{3.4 Interruption, etc.}  There are still no worker threads, but a potential strategy has been devised for them, and the Ares {\sc pma} should support parallel execution.  There are still no event timeouts, but we do not expect them to be difficult to implement.

\paragraph{3.5 Solid-State Interpreter}  While there is not yet a unikernel, a specification has been written for one.  There is no Raft cluster but the \lstinline[style=inlinecode]{%phoenix} app now uses encrypted remote scry to back up application state.  There have also been proposals to use FoundationDB to replicate event log storage.

\paragraph{4.2 Larval State}

@@

\paragraph{5.1 Nouns}  The Murmur3 system was aspirational but has been subsequently realized.  A few years ago, a horrendous bug was found and fixed in the ``mug'' short hash for cells:  the algorithm was symmetric, so \lstinline[style=inlinecode]{[1 0]} hashed to the same value as \lstinline[style=inlinecode]{[0 1]}.  This led to far more hash collisions than there should have been, degrading the asymptotics of a number of data structures.

\paragraph{6.3 Nock Optimization}  The ``jet dashboard'' (the module in the runtime responsible for registering, matching, and running jets) has been rewritten multiple times.  \patp{fodwyt-ragful} has proposed a jet dashboard that would match Nock formulas on the basis of noun equality rather than matching Nock cores (a code/data pair used as the unit of code organization) while ignoring mutable slots, the way the previous ones have worked.  This would have the advantage of simplicity, but it could be more difficult to implement more complex pieces of code, such as closures.

The Ares project has another new approach to jet matching, which is to resolve the matching statically during a transpilation step from Nock to a more optimized representation.  This is facilitated by a novel static analysis technique for Nock, called Subject Knowledge Analysis ({\sc ska}), which analyzes the data dependencies within a Nock formula, allowing inference about the tree shape of the ``subject'' (environment, or scope) at different points of the computation.  {\sc ska} should allow for more efficient registerization of intermediate values, and will also provide enough information to statically compute whether a jet will match.  The goal is for Ares to convert a hot loop in Nock into similar machine code that would be emitted by a C compiler, including direct jumps back to the start of the loop.

The Rust interpreter project mentioned in the whitepaper is dead, but in the meantime several other interpreters have been written:

\begin{enumerate}
  \item  Jaque, written using the Graal/Truffle {\sc jit} framework;
  \item  NockJS, an interpreter in JavaScript that uses trampolining;
  \item  \texttt{cl-urbit}, a Common Lisp interpreter used as a jet dashboard testbed; and
  \item  Ares, written in Rust and C, intended as the next mainline interpreter version.
\end{enumerate}

@@

\paragraph{6.5 Transition Function}

The whitepaper mentions off-handedly that Nock could not be replaced.  While this author does not expect Nock to be changed, much less replaced, this could be done if needed.  Such implementation is left as an exercise for the reader.

\paragraph{7.1 Hoon Semantics}
